---
title: "ACE rare variant burden tutorial: Sample-level QC and cohort characterization"
format:
  html:
    toc: true
    fig-width: 10
    fig-height: 7
---

# 1 Introduction

In any genetic analysis, it is foundational to ensure that the samples included in the analysis are of high quality and analysis-ready (standardized, cleaned, integrated). As a researcher, it is essential to understand your data and the characteristics of your cohort as well as data distributions before you begin to design how you will approach it.

This tutorial will begin by conducting sample-level quality control (QC) and sample characterization.

## 1.1 Set up and load libraries

R uses libraries which are sets of functions.

```{r, warn=FALSE, message=FALSE}
library(tidyverse)
# Tidyverse is a package collection of dplyr, stringr, purrr, ggplot2, readr, tidyr, tibble, forcats. If you do not have tidyverse installed and want to save a bit of time, look for `library(dplyr)`, `library(stringr)`, `library(readr)`, and `library(ggplot2)`. 

library(patchwork)
# This package is used for graphical layouts

library(ggvenn)
# This package is used to make Venn diagrams

library(skimr)
# This package allows quick summarization of data tables
```

If you need to install any of these packages, you can use the sidebar tab <Packages>, press <Install>, and type the name to get the package from CRAN (Comprehensive R Archive Network).
You can do the same thing by typing: install.packages("SKAT") and pressing enter.
If an R package is not aviailable through CRAN, it may be available through Bioconductor. Run:

`if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("VariantAnnotation")`


# 2 Load data files into the environment

Begin by loading data files from the Uganda ACE server (path: /etc/ace-data/home/jberghout/2026_ACE_bioinfo_burden/data) or Github (https://github.com/jberghout/ACE_rare_burden/tree/main/data) to the local environment.

The paths here are for the Uganda HPC. If you have downloaded the files, edit the path so that this code knows where to find them. For example: `read_delim(~/Downloads/participant_data.tsv`, ...)

```{r}
#Load file with sample data
participant_status <- read_delim("2026_ACE_bioinfo_burden/data/participant_data.tsv", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)
```

```{r}
#Load file with sample level QC
sample_qc_flags <- read_delim("2026_ACE_bioinfo_burden/data//sample_qc_flags.tsv", 
    delim = "\t", escape_double = FALSE, trim_ws = TRUE)
```

```{r}
#Load ancestry file
ancestry_table <- read_delim("2026_ACE_bioinfo_burden/data/ancestry_table.tsv", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)
```

## 2.1 Look at the data

How many samples are in the data file? What columns and data does it contain?

```{r}
participant_status %>%
  summarize(n_distinct(sample_id))
```
```{r}
skim(participant_status)
```

We see three phenotype-associated columns, review the values and distributions of each

```{r}
# Count cases and controls
participant_status %>%
  group_by(status) %>%
  summarize(n=n_distinct(sample_id)) 

# Fraction male and female by group
participant_status %>%
  group_by(status, sex) %>%
  summarize(n=n_distinct(sample_id)) %>%
  group_by(status) %>%
  mutate(percentage = 100 * n / sum(n)) %>%
  ungroup() %>%
  arrange(status, desc(n))

# Plot respiratory infections by case/control status
participant_status %>%
  ggplot(aes(x = status, fill = respiratory_infection)) +
  geom_bar(position = "fill", color = "white") +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = c("no" = "slateblue", "yes" = "red")) +
  labs(
    title = "Fraction with Respiratory Infections",
    x = "Affected",
    y = "Fraction of samples",
    fill = "Respiratory infections")

# Plot pancreatitis by case/control status
participant_status %>%
  ggplot(aes(x = status, fill = pancreatitis)) +
  geom_bar(position = "fill", color = "white") +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = c("no" = "slateblue", "yes" = "red")) +
  labs(
    title = "Fraction with Pancreatitis",
    x = "Affected",
    y = "Fraction of samples",
    fill = "Pancreatitis")

# Plot heart rate by case/control status
participant_status %>%
  ggplot(aes(x = status, y = heart_rate, fill = status)) +
  geom_boxplot() +
  scale_fill_manual(values = c("control" = "slateblue", "case" = "red")) +
  labs(
    title = "Heart Rate by Group",
    y = "Heart Rate",
    fill = "Group")
```

## 2.2 Compare the data file to the qc file and ancestry files

We will be joining these files together by using the sample_id column as a key. Before we begin, we want to quickly confirm that all three tables carry the same sample IDs so we do not have missing data at the end.

Are there any samples in the data files that are not in the qc file?

```{r}
# Count the number of samples in the data file that are not in the quality control (QC) file
setdiff(participant_status$sample_id, sample_qc_flags$sample_id)

# Plot a venn diagram of the overlap
data_vs_qc <- list(
  QC = sample_qc_flags$sample_id,
  data = participant_status$sample_id
)
ggvenn(data_vs_qc)

# What about the ancestry files?
data_vs_ancestry<-list(
  ancestry = ancestry_table$`#sample_id`,
  data = participant_status$sample_id
)
ggvenn(data_vs_ancestry)
```

# 3. Use the Quality Control file to filter the analysis-ready sample set

We only want to do analysis on samples that are acceptably high-qualiy. Incorporating low-quality samples can introduce noise and bias into the analysis, leading to either or both false positives (thinking we find a signal but it is not true) and false negatives (failing to detect a true signal).

To ensure that we are only including high-quality samples, we will use these filters:

1.  All included samples have a high call rate, aka, good coverage of the genome.
    1.  This means the proportion of variants that were successfully genotyped is high and missingness is low.
    2.  Low call rate often means that the DNA input was fragmented or poor quality, that the sequencing read depth was too low to be confident, or that there were problems aligning the raw data to the genome properly.
2.  Samples are not outliers for technical metrics or biological expectations.
3.  The samples each have a genetic sex that is concordant with the metadata reported sex
    1.  This guards against samples being mislabeled or swapped
4.  The samples have a clear documented phenotype that we can use in this analysis; not ambiguous or missing

## 3.1. Call rate/missingness threshold

Thresholds can be set using field-standard values (like \>95% call rate) or using a data-driven approach that reflects the distribution of our cohort. We'll use one of each below, starting with a simple threshold for call rate.

```{r}
# Call rate threshold; set to 95% as a minimum. This is a field-standard choice.
call_rate_threshold <- 0.95
```

Apply the 95% call rate threshold

```{r}
# Add a new column which indicates whether the sample passes or fails this filter
sample_qc_flags <- sample_qc_flags %>%
  mutate(pass_call_rate = call_rate >= call_rate_threshold)

# See how many samples pass and fail this filter
sample_qc_flags %>%
  group_by(pass_call_rate) %>%
  summarize(n = n_distinct(sample_id)) %>%
  arrange(desc(n))
```

Three failing samples is a reasonable (small) fraction, so we will apply this filter in step 3.3. If the number of failing samples was very large, we would want to investigate why.

## 3.2 Heterozygosity filter

Some kinds of genetic values and rates can vary according to the population(s) represented in the cohort and should be selected accordingly. For an example of this, we will use a data-driven approach to define the thresholds for average heterozygosity where outliers to be removed are either +/- 3 standard deviations from the cohort average heterozygosity rate.

Numbers generated in this tutorial reflect simulated data and may not be appropriate to use in a real analysis.

### 3.2.1 Define a threshold from the data

```{r}
# First, plot the distribution of this value before we apply the thresholds to see what our choices will do
qc_stats<-sample_qc_flags %>%
  summarize(
    mu = mean(het_rate, na.rm = TRUE),
    sd = sd(het_rate, na.rm = TRUE)
  ) %>%
  mutate(
    lo = mu - 3 * sd,
    hi = mu + 3 * sd
  )

sample_qc_flags %>%
  ggplot(aes(x = het_rate)) +
  geom_histogram(bins = 40, fill = "grey75", color = "white") +
  geom_vline(xintercept = qc_stats$mu, color = "red", linewidth = 1) +
  geom_vline(
    xintercept = c(qc_stats$lo, qc_stats$hi),
    color = "blue",
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    title = "Histogram of het_rate with Mean and ±3 SD",
    x = "het_rate",
    y = "Count"
  ) +
  theme_minimal()

# Set the thresholds based on +/- 3 standard deviations from the mean after reviewing the data to ensure these choices are acceptable for this dataset.
het_mean <- mean(sample_qc_flags$het_rate)
het_sd   <- sd(sample_qc_flags$het_rate)
HET_LOW  <- het_mean - 3 * het_sd
HET_HIGH <- het_mean + 3 * het_sd
```

## 3.2.2 Use the heterozygosity filter and see how many samples pass and fail this filter

We will first do this by creating a new variable that flags samples as passing or failing the filter so we can ensure the thresholds are reasonable and we can easily trace how many and which samples are being removed.

```{r}
# Create a new variable to flag passing and failing samples based on the heterozygosity thresholds
sample_qc_flags <- sample_qc_flags %>%
  mutate(het_z = scale(het_rate)[,1]) %>% 
  mutate(pass_het = het_rate >= HET_LOW & het_rate <= HET_HIGH)

# Count the number of samples that pass and fail this filter
sample_qc_flags %>%
  group_by(pass_het) %>%
  summarize(n = n_distinct(sample_id)) %>%
  arrange(desc(n))
```

## 3.3 Apply the technical QC filters to the sample metadata file

There are several ways to apply this filter easily. Here we construct the code to create two new columns, one that reads TRUE/FALSE for whether the sample passed the call rate QC filter in the sample_qc_flags datasheet and one that reads TRUE/FALSE for whether the sample passed the heterozygosity QC filter.

```{r}
participant_status <- participant_status %>%
  mutate(passed_call_rate_qc = 
      sample_id %in% sample_qc_flags$sample_id[sample_qc_flags$pass_call_rate == TRUE]) %>%
  mutate(passed_het_qc = 
      sample_id %in% sample_qc_flags$sample_id[sample_qc_flags$pass_het == TRUE])

# Count passing and failing samples for each filter
participant_status %>%
  group_by(passed_call_rate_qc, passed_het_qc) %>%
  summarize(n = n_distinct(sample_id)) %>%
  arrange(desc(n))
```

# 4. Sex check

In real data, the genetic sex of a DNA sample is determined by looking for heterozygosity on the X chromosome and/or presence of SNPs that map to the Y-chromosome (not in the pseudoautosomal region) which can be done with peddy, somalier, or PLINK. In this tutorial, discordant sex is provided in the qc_flag column.

# 4.1 Review the QC flags to identify any samples with discordant sex

Determine which values are present in the column `qc_flag`

```{r}
unique(sample_qc_flags$qc_flag)
```

Count and take the opportunity to review all available data. In real data, it may be useful to return to original sample files for any discrepant samples and ensure there are no typos or errors.

```{r}
sample_qc_flags %>%
  group_by(qc_flag) %>%
  summarize(n = n_distinct(sample_id),
            samples = head(paste(unique(sample_id), collapse = ", "), 5)) %>%
  arrange(desc(n))
```

# 4.2 Flag samples with discordant sex in the metadata

As previously, create a new column that flags our participant data table with TRUE/FALSE for whether the sample has discordant sex.

```{r}
participant_status <- participant_status %>%
  mutate(discordant_sex = sample_id %in% sample_qc_flags$sample_id[sample_qc_flags$qc_flag == "sex_discordance"])
```

# 5. Use sample QC data to finalize the analysis-ready sample set

After annotating the quality control failures onto the participant status table, we now want to remove all the samples that failed to create our final analysis-ready sample set.

```{r}
participant_final<-participant_status %>%
  filter_out(discordant_sex == TRUE) %>%
  filter_out(passed_call_rate_qc == FALSE) %>%
  filter_out(passed_het_qc == FALSE)
```

# 6. Characterize the cohort

Try a few on your own:

\- How many samples?

\- How many cases and controls?

# 7. Incorporate ancestry data and principal components (PCs) into the participant_final table

In genetic analyses, population structure and relatedness are major potential confounder to accurately discovering true gene-to-trait associations. Samples that are more closely related to each other due to a more recent common ancestry or shared evolutionary history may share many unique genetic variants and patterns that are not related to the trait of interest.

To account for this, we will incorporate principal components (PCs) into the participant_final table.

## 7.1 Review the ancestry data

The ancestry table (here, generated by somalier) contains the predicted ancestry for each sample and the first 10 PCs. The predicted ancestry is a categorical variable that indicates the most likely ancestry of each sample based on the genetic data. The PCs are continuous variables that capture the major axes of genetic variation in the cohort.

We will use these ancestry PCs as covariates to reduce confounding from population stratification, helping separate true genetic signals from ancestry-driven differences.

```{r}
# Notice that the ancestry table also includes some reference samples from the 1000 Genomes Project in addition to our samples
head(ancestry_table)
skim(ancestry_table)
```

Plot the first two PCs to visualize.

```{r}
ancestry_table %>%
  ggplot(aes(x = PC1, y = PC2, color = predicted_ancestry)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Ancestry Distribution", x = "PC1", y = "PC2")
```

## 7.2 Join the ancestry data to the participant_final table

Join the ancestry data to the participant_final table by sample_id. This will add the predicted ancestry and the first 10 PCs to the participant_final table. Using a `left_join()` ensures that we only use our samples and drop the reference samples.

In this table join, you will notice that we need to join the ancestry table by the column `#sample_id` and the participant_final table by the column `sample_id`. This is because the ancestry table generated by Somalier uses a `#` prefix in the column name. R uses the `by` argument in the `left_join()` function to specify how to join these two tables together.

```{r}
participant_final <- participant_final %>%
  left_join(ancestry_table %>% 
            dplyr::select(`#sample_id`, predicted_ancestry, PC1, PC2, PC3, PC4, PC5, PC6, PC7, PC8, PC9, PC10), 
            by = c("sample_id" = "#sample_id"))
```

# 7.3 Plot PCA of the cohort colored by case/control status

```{r}
# PC1 vs PC2 (the most informative PCs)
participant_final %>%
  ggplot(aes(x = PC1, y = PC2, color = status)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Ancestry Distribution of the Cohort", x = "PC1", y = "PC2")
```

```{r}
# PC1 vs PC3 going down the list...
participant_final %>%
  ggplot(aes(x = PC1, y = PC3, color = status)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Ancestry Distribution of the Cohort", x = "PC1", y = "PC2")
```

# Before the next portion:

Save the participant_final table as a tsv in your results folder.

If you continue working in the same active session and can see participant_final in your sidebar 'Environment', you do not need to save the file and re-load the file, so this step is optional during the tutorial. While working with your real data in the future, you will want to save your `final` dataset for archiving purposes.
 
```{r, eval = FALSE}
write_tsv(participant_final, "2026_ACE_bioinfo_burden/results/participant_final.tsv")
```
